The Weakly Nonlinear Magnetorotational Instability

### 2016-08-09 paper/thingap_mri_paper.tex
run helical mri critical param run w new complex eigenvalue finder

We note several other important analytical studies of MRI saturation. \citeauthor{Knobloch:2005ba} (\citeyear{Knobloch:2005ba}) analyze the MRI in the strongly nonlinear regime, by following the already-developed MRI modes into asymptotic saturation. They find that this saturation proceeds via modification of the background shear. \citeauthor{Knobloch:2005ba} consider a thin-gap regime as well, and so their theory may be considered the strongly nonlinear analogue to the one developed here. \citei{Vasil:2015} examines the weakly nonlinear MRI in a thin-gap regime from a heuristic perspective, finding intriguing similarities between the MRI system and an elastodynamic instability. Several authors have investigated the behavior of the MRI when the boundary conditions are shear periodic, and so the MRI has no mechanism by which to modify the background shear flow profile. In this paradigm linear MRI growth is dominated by channel modes, a type of MRI mode that, for periodic boundary conditions, are exact solutions of both the linear and nonlinear MRI equations (\citei{Goodman:1994ul}). In this regime the MRI saturates via parasitic instabilities, which feed off and destroy the primary MRI modes. Analytical investigation of this case reveals that MRI saturation can be caused by parasitic Kelvin-Helmholtz and tearing mode instabilities, depending on parameter regime (\citei{Pessah:2010ic}). The theory of MRI channel mode parasites is relatively robust (see also \citei{Pessah:2009gm}, \citei{Latter:2010iz}, \citei{Rembiasz:2016}), but they are mostly an artifact of the local approximation, and not germane to global analyses like the one presented here (see \citei{Latter:2015} for a detailed analysis of the relation between local and global MRI modes).

>>>>

We note several other important analytical studies of MRI saturation. \citeauthor{Knobloch:2005ba} (\citeyear{Knobloch:2005ba}) analyze the MRI in the strongly nonlinear regime, by following the already-developed MRI modes into asymptotic saturation. They find that this saturation proceeds via modification of the background shear. \citeauthor{Knobloch:2005ba} consider a thin-gap regime as well, and so their theory may be considered the strongly nonlinear analogue to the one developed here. \citei{Vasil:2015} examines the weakly nonlinear MRI in a thin-gap regime from a heuristic perspective, finding deep mathematical similarities between the MRI system and an elastodynamic instability. Several authors have investigated the behavior of the MRI when the boundary conditions are shear periodic, and so the MRI has no mechanism by which to modify the background shear flow profile. In this approximation linear MRI growth is dominated by channel modes, a type of MRI mode that, for periodic boundary conditions, are exact solutions of both the linear and nonlinear MRI equations (\citei{Goodman:1994ul}). In this regime the MRI saturates via parasitic instabilities, which feed off and destroy the primary MRI modes. Analytical investigation of this case reveals that MRI saturation can be caused by parasitic Kelvin-Helmholtz and tearing mode instabilities, depending on parameter regime (\citei{Pessah:2010ic}). The theory of MRI channel mode parasites is relatively robust (see also \citei{Pessah:2009gm}, \citei{Latter:2010iz}, \citei{Rembiasz:2016}), but they are mostly an artifact of the local approximation, and not germane to global analyses like the one presented here (see \citei{Latter:2015} for a detailed analysis of the relation between local and global MRI modes).

### 2016-08-09 paper/widegap_mri_paper.tex
run helical mri critical param run w new complex eigenvalue finder

Although the MRI is broadly important to many astrophysical systems, many of its general properties, and in particular its nonlinear saturation mechanism, remain poorly understood. The diversity of astrophysical systems which may permit the MRI admits an enormous parameter space to be explored. In protoplanetary disks, for example, the behavior and evolution of the MRI may change drastically depending on the properties of the magnetic field, the disk composition, disk geometry, etc. Numerical simulation of realistic disk physics is currently an area of intense focus, and is enabling the study of nonideal MHD effects, disk stratification, nonequilibrium chemistry, and other complex physics that does not lend itself easily to analytic study. Still, computational costs inevitably constrain numerical approaches. MRI saturation is a complicated nonlinear problem which may depend on the assumptions and approximations adopted by simulations in nonobvious ways. Analytic methods can play a powerful role in elucidating the mechanisms responsible for MRI saturation. For instance, analytical approaches have revealed the mechanism that likely governs saturation in the ``shearing box" approximation. The shearing box is an oft-invoked local approximation in which a section of a disk is represented by solving the MHD equations in an isolated region subject to shear periodic boundary conditions. The shearing box is an inexpensive computational framework, and has been widely used to study many aspects of the MRI (). However, while the MRI is a local instability, certain properties of the local problem are not generic to the global problem. In the shearing box, linear evolution is dominated by a class of MRI mode known as channel modes. These are linear modes which also happen to be exact solutions to the \textit{nonlinear} local MRI equations. Runaway growth is avoided in this paradigm by the instability of the channel modes themselves, which are destroyed by parasitic (secondary) instabilities (\citei{Goodman:1994ul}, \citei{Pessah:2010ic}). The growth of parasitic modes provides a saturation avenue for channel mode-dominated flows, yet this is unlikely to be the dominant saturation mechanism in laboratory experiments or astrophysical disks, as channel modes are artificially over-prominent in the shearing box (e.g. \citei{Latter:2015}). Thus while the shearing box may accurately approximate many features of the global MRI, the saturation mechanism is not among them.

@@

The theory we develop here employs realistic radial boundary conditions, and so MRI modes are global and channel modes (at least in their purest, radially independent are not present. A number of saturation mechanisms have been proposed for the MRI which do not rely on channel modes dominating the flow. The MRI feeds off of the free energy from differential rotation, and so a modification of the background shear may cause saturation (e.g. \citei{Knobloch:2005ba}, \citei{Umurhan:2007hs}). The MRI may transfer its free energy into the magnetic field, and saturate when the field is too strong to be susceptible to the MRI (e.g. \citei{Ebrahimi:2009ey}). The MRI may saturate differently depending on the particular parameter regime under investigation, and so our challenge is not only in identifying possible saturation mechanisms, but in understanding how and when each applies in different astrophysical environments.

@@

@@

\section{Channel Modes}
Channel modes are radially independent linear MRI modes which are, under certain conditions, exact nonlinear solutions to the MRI equations.

@@

@@

@@

>>>>

Although the MRI is broadly important to many astrophysical systems, many of its general properties, and in particular its nonlinear saturation mechanism, remain poorly understood. The diversity of astrophysical systems which may permit the MRI admits an enormous parameter space to be explored. In protoplanetary disks, for example, the behavior and evolution of the MRI may change drastically depending on the properties of the magnetic field, the disk composition, disk geometry, and so forth. Numerical simulation of realistic disk physics is currently an area of intense focus, and is enabling the study of nonideal MHD effects, disk stratification, nonequilibrium chemistry, and other complex physics that does not lend itself easily to analytic study (\citei{Fleming:2003fs}, \citei{Bai:2011cm}, \citei{Flock:2013}, \citei{Suzuki:2014vh}, to name only a few). Still, computational costs inevitably constrain numerical approaches. MRI saturation is a complicated nonlinear problem which may depend on the assumptions and approximations adopted by simulations in nonobvious ways. Analytic methods can play a powerful role in elucidating the mechanisms responsible for MRI saturation. For instance, analytical approaches have revealed the mechanism that likely governs saturation in the ``shearing box" approximation. The shearing box is an oft-invoked local approximation in which a section of a disk is represented by solving the MHD equations in an isolated region subject to shear periodic boundary conditions. The shearing box is an inexpensive computational framework, and has been widely used to study many aspects of the MRI (). However, while the MRI is a local instability, certain properties of the local problem are not generic to the global problem. In the shearing box, linear evolution is dominated by a class of MRI mode known as channel modes. These are linear modes which also happen to be exact solutions to the \textit{nonlinear} local MRI equations. Runaway growth is avoided in this paradigm by the instability of the channel modes themselves, which are destroyed by parasitic (secondary) instabilities (\citei{Goodman:1994ul}, \citei{Pessah:2010ic}). The growth of parasitic modes provides a saturation avenue for channel mode-dominated flows, yet this is unlikely to be the dominant saturation mechanism in laboratory experiments or astrophysical disks, as channel modes are artificially over-prominent in the shearing box (e.g. \citei{Latter:2015}). Thus while the shearing box may accurately approximate many features of the global MRI, the saturation mechanism is not among them.

@@

The theory we develop here employs realistic radial boundary conditions, and so MRI modes are global and channel modes are not present. A number of saturation mechanisms have been proposed for the MRI which do not rely on channel modes dominating the flow. The MRI feeds off of the free energy from differential rotation, and so a modification of the background shear may cause saturation (e.g. \citei{Knobloch:2005ba}, \citei{Umurhan:2007hs}). The MRI may transfer its free energy into the magnetic field, and saturate when the field is too strong to be susceptible to the MRI (e.g. \citei{Ebrahimi:2009ey}). The MRI may saturate differently depending on the particular parameter regime under investigation, and so our challenge is not only in identifying possible saturation mechanisms, but in understanding how and when each applies in different astrophysical environments.

@@

\section{Results and Discussion}

\subsection{Standard MRI}

@@

%\section{Channel Modes}
%Channel modes are radially independent linear MRI modes which are, under certain conditions, exact nonlinear solutions to the MRI equations.

\subsection{Helical MRI}

When $\xi$ in Equation \ref{eq:unperturbed_matrix_eqns} is not equal to zero, the helical MRI arises. We examine a single fiducial helical MRI case, for the parameters listed in Table \ref{table:parameters}. At the conclusion of the weakly nonlinear analysis, we find that the coefficients of Equation \ref{eq:gle} are $b = $, $d = $, and $c = $, i.e. complex. The marginal helical MRI is thus described by a complex GLE.

@@

\bibitem[{Bai(2011)}]{Bai:2011cm}
Bai, X. 2011, ApJ 739, 50

@@

\bibitem[{Fleming \& Stone(2003)}]{Fleming:2003fs}
Fleming, T. \& Stone, J. 2003, ApJ 585, 908

\bibitem[{Flock et al.(2013)}]{Flock:2013}
Flock, M., Fromang, S., Gonz\'alez, M., Commer\c{c}on, B. A\&A 560, A43

@@

\bibitem[{Suzuki \& Inutsuka(2014)}]{Suzuki:2014vh}
Suzuki, T.K. \& Inutsuka, S. 2014, ApJ 784, 121
